A data-driven model of nucleosynthesis with chemical tagging in latent space

### 2019-06-20 article/mcfa.tex
things

\newcommand{\Gaia}{\project{Gaia}}
\newcommand{\gaia}{\project{gaia}}
\newcommand{\Galah}{\project{Galah}}
\newcommand{\GALAH}{\project{GALAH}}
\newcommand{\APOGEE}{\project{APOGEE}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}

@@

\newcommand{\NumData}{N}
\newcommand{\NumDimensions}{D}
\newcommand{\numdata}{n}
\newcommand{\numdimensions}{d}
\newcommand{\NumLatentFactors}{J}
\newcommand{\numlatentfactors}{j}
\newcommand{\NumComponents}{K}
\newcommand{\numcomponents}{k}
\newcommand{\likelihood}{\mathcal{L}}
\newcommand{\ExperimentHash}{8761b}

@@

%[0000-0003-0174-0564]
\author{Some combination (TBD) of: Andrew R. Casey}
\affiliation{School of Physics \& Astronomy, 
			 Monash University,
			 Wellington Rd, Clayton 3800, Victoria, Australia}
\affiliation{Faculty of Information Technology, 
			 Monash University, 
			 Wellington Rd, Clayton 3800, Victoria, Australia}

\author{John Lattanzio}
\affiliation{School of Physics \& Astronomy, 
			 Monash University,
			 Wellington Rd, Clayton 3800, Victoria, Australia}

\author{Aldeida Aleti}
\affiliation{Faculty of Information Technology, 
			 Monash University, 
			 Wellington Rd, Clayton 3800, Victoria, Australia}

\author{David Dowe}
\affiliation{Faculty of Information Technology, 
			 Monash University, 
			 Wellington Rd, Clayton 3800, Victoria, Australia}

@@

\begin{abstract}
Chemical tagging seeks to identify unique star formation sites from
present-day stellar abundances.
Previous techniques have treated each abundance dimension as being
statistically independent, despite theoretical expectations that
many elements can be produced by more than one nucleosynthetic process.
In this work we introduce a data-driven model of nucleosynthesis 
where a set of latent factors (e.g., nucleosynthetic yields) contribute
to all stars with different scores, and clustering (e.g., chemical tagging) 
is modelled by a mixture of multivariate gaussians
in lower-dimensional latent space.
We use an exact method to simultaneously estimate the factor scores for
each star, the partial assignment of each star
to each cluster, and the latent factors common to all stars.
We use an information-theoretic Bayesian principle to estimate the number of
latent factors and clusters.
We find that five latent factors are preferred to explain 18 chemical elements
in the second \Galah\ data release.
From these latent factors we identify the rapid- and slow-neutron capture 
processes, as well as latent factors consistent with Fe-peak and 
$\alpha$-element production, and a fifth latent factor where only K and Zn
dominate. We discuss examples where the inferred latent factors deviate from 
expected yields, which includes an  indication that Sc is associated with the r-process.
We find three clusters are preferred given 1,072 stars in \Galah,
two of which are representative of the [$\alpha$/Fe]-poor and 
[$\alpha$/Fe]-rich populations in the Milky Way disk. 
We release accompanying software that scales well with the available data,
allowing for model parameters to be optimised in seconds given a fixed number
of latent factors, components, and $\sim10^6$ abundance measurements.
\end{abstract}

\keywords{methods: statistical}

@@

%Alternative approaches are perhaps more challenging, as i
It is not trivial to confidently estimate
the nucleosynthetic yields that have contributed to the chemical abundances of each star. There are
qualitative statements that can be made for large numbers of stars, or particular
types of stars, but quantifying the precise contribution of different processes
to each star is an unsolved problem. For example, the so-called [$\alpha$/Fe] `knee' in
abundance ratios in the Milky Way can qualitatively be explained by 
core-collapse supernovae being the predominant nucleosynthetic process in the
early Milky Way before Type Ia supernovae made a significant contribution, but 
efforts to date have not sought to try to explain the detailed abundances of a 
single star as a contribution of yields from different systems (e.g., 20\% of the
Ba was produced by 3 M$_\odot$ asymptotic giant branch stars). This is in part 
because of the challenging and degenerate nature of the problem as described, 
and is complicated by the differences in yield predictions that account from 
prescriptions used in different theoretical models.

@@

\noindent{}where $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$ 
matrix of factor loads that is common to all data points, $\NumLatentFactors$ is
the number of latent factors, and the factor scores 
for the $\numdata$th data point
\begin{equation}
	\factorscores_\numdata \sim \mathcal{N}(\vec\xi_\numcomponents, \vec\Omega_\numcomponents)
\end{equation}
\noindent{}are drawn from\footnote{For clarifying nomenclature across disciplines, the terminology e$z \sim \mathcal{N}(0, 1)$ indicates that the variable $z$ is drawn from a standard normal distribution.} the $\numcomponents$th multivariate normal distribution.
The mean vector $\vec\mu$ describes the mean datum in each dimension.
The factor scores for all data points $\factorscores$ is then a 
$\NumData \times \NumLatentFactors$ matrix, where each data point has a partial
association to the components in latent space. 
We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \textrm{diag}(\specificvariance)\right)$
is independent of the latent space, and $\specificvariance$ is a
vector of variances in each $\NumDimensions$ abundance dimensions.
In this model each data point can be represented as being drawn
from a mixture of multivariate normal components, except the components
are \emph{clustered in the latent space} $\factorscores$ and projected
into the data space by the factor loads $\factorloads$.

@@

%Throughout this work we found that the optimised model parameters appear
%to be insensitive to the choice of initialisation up until orthogonal
%rotation of the latent space.

@@

@@

We assume uniform priors for the component means in latent space $\scoremeans$, 
and a conjugate inverted Wishart prior for the component covariance matrices
$\scorecovs$ \citep[Section 5.2.3;][]{Knorr-Held:2000},
\begin{equation}
	p(\scoremeans_\numcomponents,\scorecovs_\numcomponents) \propto |\scorecovs_\numcomponents|^{\frac{1}{2}(\NumLatentFactors + 1)} \quad .
\end{equation}
We approximate the determinate of the Fisher information of a multivariate normal $|\mathcal{F}(\scoremeans,\scorecovs)|$
as $|\mathcal{F}(\scoremeans)||\mathcal{F}(\scorecovs)|$ \citep{Oliver:1996,Figueiredo:2002} where
\begin{equation}
	|\mathcal{F}(\scoremeans)| = (\NumData\weight_k)^\NumLatentFactors|\scorecovs_k|^{-1}
\end{equation}
\begin{equation}
	|\mathcal{F}(\scorecovs)| = (\NumData\weight_k)^{\frac{1}{2}\NumLatentFactors(\NumLatentFactors+1)}2^{-\NumLatentFactors}|\scorecovs_k|^{-(\NumData+1)}
\end{equation}
\noindent{}such that 
% todo put line in equation.
\begin{eqnarray}
	I(\scoremeans,\scorecovs) &=& -\sum_{\numcomponents=1}^{\NumComponents}\log{p(\scoremeans_k,\scorecovs_k)} + \frac{1}{2}\sum_{\numcomponents=1}^{\NumComponents}\log{|\mathcal{F}(\scoremeans_k,\scorecovs_k)|} \nonumber \\
	&=& \frac{1}{2}\sum_{\numcomponents=1}^{\NumComponents}\log\left[(\NumData\weight_k)^{\frac{1}{2}\NumLatentFactors(\NumLatentFactors+3)}2^{-\NumData}|\scorecovs_k|^{-(\NumData + 2)}\right] \nonumber \\
	&& \cdots \,\, -\sum_{\numcomponents=1}^{\NumComponents}\log{|\scorecovs_k|}^{\frac{1}{2}(\NumLatentFactors + 1)} \\
I(\scoremeans,\scorecovs) &=& \frac{1}{4}\NumLatentFactors(\NumLatentFactors+3)\sum_{\numcomponents=1}^\NumComponents\log{\NumData\weight_k} - \frac{KD}{2}\log{2} \nonumber \\ 
&& \cdots \,\, -\frac{1}{2}(2\NumLatentFactors+3)\sum_{k=1}^{K}\log{|\scorecovs_k|}  \quad . \label{eq:prior_xi_omega} 
\end{eqnarray}

@@

The Fisher information for $\factorloads$ \todo{needs to be double checked.}
Thus the message length to encode $\factorloads$ is given by
\begin{eqnarray}
	I(\factorloads) &=& -\log\left(\frac{p(\factorloads)}{\sqrt{|\mathcal{F}(\factorloads)|}}\right) \nonumber \\
I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} \nonumber \\
	&& \cdots \,\, + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\
	&& \cdots \todo{+\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \quad . \label{eq:prior_L}
\end{eqnarray}

Combining equations \ref{eq:prior_J}, \ref{eq:prior_K}, \ref{eq:prior_pi}, \ref{eq:prior_xi_omega}, and \ref{eq:prior_L} with equation \ref{eq:mml} leads to the full message length:
\begin{widetext}
\begin{eqnarray}
	I(\vec\Psi, \vec\data) &=& -\log\likelihood(\vec\data|\vec\Psi)
 +\frac{1}{4}\left(\NumLatentFactors + 4\right)\left(\NumLatentFactors - 1\right)\sum_{\numcomponents=1}^\NumComponents\log\weight_\numcomponents + \left(\NumComponents - \frac{1}{2}\right)\log{\NumData}
 +\frac{1}{2}\NumDimensions\log|\textrm{Cov}\left(\factorloads\transpose\right)| \nonumber \\
  && \cdots -\frac{1}{2}\left(D-J-1\right)\log|\factorloads\transpose\factorloads| + \textrm{Tr}\left(\textrm{Cov}\left(\factorloads\transpose\right)^{-1}\factorloads\transpose\factorloads\right) - \left(\NumLatentFactors + \frac{3}{2}\right)\sum_{\numcomponents=1}^\NumComponents\log|\scorecovs_\numcomponents|  -\log\Gamma\left(\NumComponents\right) - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\
&& \cdots +\frac{Q}{2}\log\kappa_q +\frac{1}{2}\left[\NumLatentFactors(\NumDimensions + 2) + \NumComponents(2-\NumData)\right]\log{2} + \todo{\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \quad . \label{eq:message-length}
 \end{eqnarray}
\end{widetext}

@@

\subsection{Toy model with generated data} \label{sec:exp-1}

We generated a data set with ${\NumData = 1}$00,000 data points, each with
$\NumDimensions = 15$ dimensions. We adopted a latent dimensional space of 
$\NumLatentFactors = 5$ factor loads such that the vector $\factorloads$ has
shape $\NumLatentFactors \times \NumDimensions$,
with $\NumComponents = 20$ clusters in the latent space. 
We generated the factor loads in the same way that we initialise the optimisation (Section~\ref{sec:initialisation}). The relative weights $\vec\weight$
are drawn from a multinomial distribution and the means of the clusters
in factor scores $\scoremeans$ are drawn from a standard normal
distribution. The off-diagonal entries in the covariance matrices in factor scores $\scorecovs$ are drawn from a gamma distribution $\scorecovs_{\numcomponents,i,i} \sim \vec\Gamma\left(1\right)$. The variance in 
each dimension $\specificvariance$ are also drawn $\specificvariance \sim \vec\Gamma\left(1\right)$.
The $\numdata$th data point (which belongs to the $\numcomponents$th cluster) is then
generated by drawing $\factorscores_{\numdata} \sim \mathcal{N}(\scoremeans_\numcomponents,\scorecovs_\numcomponents)$, projecting by the factor loads $\factorloads$, and adding variance $\specificvariance$.

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp1-latent.pdf}
    \caption{The inferred latent scores $\factorscores$ for our toy model
     with $\NumLatentFactors = 5$ latent factors and
		 	 $\NumComponents = 20$ components.
             Points are coloured by the inferred component.} 
    \label{fig:exp1-latent}
\end{figure*}

@@

We recorded the \emph{negative} log likelihood, the BIC, and the message length (omitting constant terms) for each permutation of $\NumLatentFactors$ and $\NumComponents$.
These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.
Unsurprisingly the negative log likelihood increases with increasing numbers of latent
factors $\NumLatentFactors$ and increasing numbers of components $\NumComponents$.
The lowest BIC value and message length is found at $\NumLatentFactors = 5$
and $\NumComponents = 20$, identical to the true values.

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp1-compare-all.pdf}
    \caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
    		 and specific variances $\specificvariance$ (right) compared to the 
		 	 true data generating values
		 	 for Experiment~1. The agreement is excellent.}
    \label{fig:exp1-compare}
\end{figure*}

It is clear from Figure~\ref{fig:experiment-1-gridsearch} that a combination of latent factors
and clustering in the latent space provides a better description of the (generated) data.
Adding components to the model does improves the log likelihood, even with a single latent factor,
but the addition of just \emph{one latent factor} improves the log likelihood more so than adding
\emph{twenty components}. Not much more can be said for this example because the true data generating process 
is drawn from a higher number of latent factors, but this toy model does illustrate how 
clustering in high dimensional data can be better described by latent factors with 
clustering in the lower dimensional latent space.
The inferred clustering in latent space is shown in Figure~\ref{fig:exp1-latent},
where each point is coloured by the inferred associated component.

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp2-\ExperimentHash-latent-factors-and-contributions-with-targets.pdf}
	\caption{Latent factors found in Experiment~2 using the \citet{Barklem:2005}
			 data. The panels on the left show the absolute entries for each factor load. 
			 On the right we show the absolute fractional contributions to
			 each element, illustrating the loads that contribute most to each element.}
    \label{fig:exp2-factor-loads}
\end{figure*}

@@

@@

In this experiment we perform blind chemical tagging using the 
photospheric abundances released as part of the second \Galah\ 
data release \citep{Buder:2018}. This data set includes
up to 23 chemical abundances reported for 342,682 stars.
Some abundances are not reported for a variety of 
reasons related to astrophysics or the adopted analyses
(e.g., a star is too hot to estimate the abundance for a given
electronic transition, or the signal-to-noise of the spectrum is too low).
Our method as described cannot efficiently handle missing data 
entries for a given star, and as such we are forced to exclude stars that
do not have a full complement of abundances. However, we also
chose to exclude some elemental abundances from our chemical
tagging experiments. For example, our experiments
with \Galah\ data do not include lithium or carbon abundances because the
photospheric values will vary throughout a star's lifetime \citep[e.g.,][]{Casey:2016b,Casey:2019}. This is true
to a small degree for many elements \citep[e.g.,][]{Dotter:2017},
but for the purposes of this experiment we assume that all other
photospheric abundances remain constant throughout a star's
lifetime.

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-latent-factors-and-contributions-with-targets.pdf}
	\caption{Latent factors inferred from the \Galah\
			 data \citep[][; thick lines]{Buder:2018}. The panels on the left show the absolute entries for each
			 factor load, where the thin lines indicate the target latent factors (see Section~\ref{sec:exp4}). On the right we show the absolute fractional contributions
			 to each element, ordered by the loads that contribute most to each
			 element.}
    \label{fig:exp3-factor-loads}
\end{figure*}

@@

\clearpage

>>>>

\newcommand{\Gaia}{\project{Gaia}}
\newcommand{\gaia}{\project{gaia}}
\newcommand{\Galah}{\project{Galah}}
\newcommand{\GALAH}{\project{GALAH}}
\newcommand{\APOGEE}{\project{APOGEE}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
\newcommand{\consider}[1]{\textcolor{blue}{#1}}

@@

\newcommand{\NumData}{N}
\newcommand{\NumDimensions}{D}
\newcommand{\numdata}{n}
\newcommand{\numdimensions}{d}
\newcommand{\NumLatentFactors}{J}
\newcommand{\numlatentfactors}{j}
\newcommand{\NumComponents}{K}
\newcommand{\numcomponents}{k}
\newcommand{\likelihood}{\mathcal{L}}
\newcommand{\ExperimentHash}{89dab}

% NEED: Final model with # of loads and # of clusters so we can:
% 		[ ] Update figure of loads (exp10-comparison)
% 		[ ] Make a clustering plot in abundances
% 		[ ] Check to see if any chemical clusters correspond to known clusters.
% 		[ ] Update discussion w.r.t. final model loads, etc.
% 		[ ] Update conclusion w.r.t. final model loads/clusters, etc, and missing data.
% 		[ ] Update abstract w.r.t. final model loads/clusters.
% 		[ ] Upload PDF to Galah survey website
%		[ ] Send email to Galah team for comment

@@

\author[0000-0003-0174-0564]{Andrew R. Casey}
\affiliation{School of Physics \& Astronomy, 
			 Monash University,
			 Wellington Rd, Clayton 3800, Victoria, Australia}
\affiliation{Faculty of Information Technology, 
			 Monash University, 
			 Wellington Rd, Clayton 3800, Victoria, Australia}

\author[0000-0003-2952-859X]{John C. Lattanzio}
\affiliation{School of Physics \& Astronomy, 
			 Monash University,
			 Wellington Rd, Clayton 3800, Victoria, Australia}

\author[0000-0002-1716-690X]{Aldeida Aleti}
\affiliation{Faculty of Information Technology, 
			 Monash University, 
			 Wellington Rd, Clayton 3800, Victoria, Australia}

\author[0000-0002-0583-5918]{David Dowe}
\affiliation{Faculty of Information Technology, 
			 Monash University, 
			 Wellington Rd, Clayton 3800, Victoria, Australia}

@@

\begin{abstract}
Chemical tagging seeks to identify unique star formation sites from
present-day stellar abundances.
Previous techniques have treated each abundance dimension as being
statistically independent, despite theoretical expectations that
many elements can be produced by more than one nucleosynthetic process.
In this work we introduce a data-driven model of nucleosynthesis 
where a set of latent factors (e.g., nucleosynthetic yields) contribute
to all stars with different scores, and clustering (e.g., chemical tagging) 
is modelled by a mixture of multivariate gaussians
in lower-dimensional latent space.
We use an exact method to simultaneously estimate the factor scores for
each star, the partial assignment of each star
to each cluster, and the latent factors common to all stars.
We use an information-theoretic Bayesian principle to estimate the number of
latent factors and clusters.
\consider{We find that five latent factors are preferred to explain 18 chemical elements
in the second \Galah\ data release.}
From these latent factors we identify the rapid- and slow-neutron capture 
processes, as well as latent factors consistent with Fe-peak and 
$\alpha$-element production, and a \consider{fifth latent factor where only K and Zn
dominate}. We discuss examples where the inferred latent factors deviate from 
expected yields, which includes an \consider{indication that Sc is associated with the r-process}.
\consider{We find three clusters are preferred given 1,072 stars in \Galah,
two of which are representative of the [$\alpha$/Fe]-poor and 
[$\alpha$/Fe]-rich populations in the Milky Way disk.}
We release accompanying software that scales well with the available data,
allowing for model parameters to be optimised in seconds given a fixed number
of latent factors, components, and $\sim10^7$ abundance measurements.
\end{abstract}

\keywords{Bayesian statistics (1900), Chemical abundances (224), Galaxy chemical evolution (580)}

@@

%Alternative approaches are perhaps more challenging, as i
It is not trivial to confidently estimate
the nucleosynthetic yields that have contributed to the chemical abundances of each star. There are
qualitative statements that can be made for large numbers of stars, or particular
types of stars, but quantifying the precise contribution of different processes
to each star is an unsolved problem. For example, the so-called [$\alpha$/Fe] `knee' in
abundance ratios in the Milky Way can qualitatively be explained by 
core-collapse supernovae being the predominant nucleosynthetic process in the
early Milky Way before Type Ia supernovae made a significant contribution, but 
efforts to date have not sought to try to explain the detailed abundances of 
stars as a contribution of yields from different systems \citep[however see][]{West:2013}.
%e.g., 20\% of the
%Ba was produced by 3 M$_\odot$ asymptotic giant branch stars).
This is in part 
because of the challenging and degenerate nature of the problem as described, 
and is complicated by the differences in yield predictions that account from 
prescriptions used in different theoretical models.

@@

\noindent{}where $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$ 
matrix of factor loads that is common to all data points, $\NumLatentFactors$ is
the number of latent factors, and the factor scores 
for the $\numdata$th data point
\begin{equation}
	\factorscores_\numdata \sim \mathcal{N}(\vec\xi_\numcomponents, \vec\Omega_\numcomponents)
\end{equation}
\noindent{}are drawn from\footnote{For clarifying nomenclature across disciplines, the terminology $z \sim \mathcal{N}(0, 1)$ indicates that the $z$ variable \emph{is drawn from} a standard normal distribution.} the $\numcomponents$th multivariate normal distribution.
The mean vector $\vec\mu$ describes the mean datum in each dimension.
The factor scores for all data points $\factorscores$ is then a 
$\NumData \times \NumLatentFactors$ matrix, where each data point has a partial
association to the components in latent space. 
We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \textrm{diag}(\specificvariance)\right)$
is independent of the latent space, and $\specificvariance$ is a
vector of variances in each $\NumDimensions$ abundance dimensions.
In this model each data point can be represented as being drawn
from a mixture of multivariate normal components, except the components
are \emph{clustered in the latent space} $\factorscores$ and projected
into the data space by the factor loads $\factorloads$.

@@

\vspace{3em}
%Throughout this work we found that the optimised model parameters appear
%to be insensitive to the choice of initialisation up until orthogonal
%rotation of the latent space.

@@

\subsection{Missing data}

The Expectation-Maximization procedure as described requires that there be no 
missing data entries in order to update our estimates of the model parameters.
In practice, however, there will often be abundance measurements that are missing
for some subset of stars. There are many potential reasons for this, including
astrophysical explanations (e.g., an absorption line was not present above the noise),
observational limitations (e.g., the signal-to-noise ratio was too low, or 
contamination by a cosmic ray), or various other reasons.

In this work we will assume that any missing data measurements are
missing at random. The missing data points can then be treated as unknown
parameters that must be solved for (and updated) at each iteration. Initially
we impute zeros for missing data entries, and at each iteration we update
this imputed value with our estimate of what the missing data values
are, given the current model parameters. Similarly, with each update we inflate
our estimates of the specific variances $\vec\psi$ based on the fraction of missing
data points in each dimension
\begin{equation}
\specificvariance\nextstep = \specificvariance\nextstep\left(\frac{N}{N-\vec{M}}\right)
\end{equation}
\noindent{}where $\vec{M}$ is a vector that contains the number of missing data entries in
all $D$ dimensions.
In Section \ref{sec:toy-model-missing-data} we show with a toy model that the latent
factor loads and scores can be reliably estimated even in the presence of high fractions
of missing data, with varying fractions of missing data in each dimension.

@@

We assume uniform priors for the component means in latent space $\scoremeans$, 
and a conjugate inverted Wishart prior for the component covariance matrices
$\scorecovs$ \citep[Section 5.2.3;][]{Knorr-Held:2000},
\begin{equation}
	p(\scoremeans_\numcomponents,\scorecovs_\numcomponents) \propto |\scorecovs_\numcomponents|^{\frac{1}{2}(\NumLatentFactors + 1)} \quad .
\end{equation}
We approximate the determinate of the Fisher information of a multivariate normal $|\mathcal{F}(\scoremeans,\scorecovs)|$
as $|\mathcal{F}(\scoremeans)||\mathcal{F}(\scorecovs)|$ \citep{Oliver:1996,Figueiredo:2002} where
\begin{equation}
	|\mathcal{F}(\scoremeans)| = (\NumData\weight_k)^\NumLatentFactors|\scorecovs_k|^{-1}
\end{equation}
\begin{equation}
	|\mathcal{F}(\scorecovs)| = (\NumData\weight_k)^{\frac{1}{2}\NumLatentFactors(\NumLatentFactors+1)}2^{-\NumLatentFactors}|\scorecovs_k|^{-(\NumData+1)}
\end{equation}
\noindent{}such that 
% todo put line in equation.
\begin{eqnarray}
	I(\scoremeans,\scorecovs) &=& -\sum_{\numcomponents=1}^{\NumComponents}\log{p(\scoremeans_k,\scorecovs_k)} + \frac{1}{2}\sum_{\numcomponents=1}^{\NumComponents}\log{|\mathcal{F}(\scoremeans_k,\scorecovs_k)|} \nonumber \\
%	&=& \frac{1}{2}\sum_{\numcomponents=1}^{\NumComponents}\log\left[(\NumData\weight_k)^{\frac{1}{2}\NumLatentFactors(\NumLatentFactors+3)}2^{-\NumData}|\scorecovs_k|^{-(\NumData + 2)}\right] \nonumber \\
%	&& \cdots \,\, -\sum_{\numcomponents=1}^{\NumComponents}\log{|\scorecovs_k|}^{\frac{1}{2}(\NumLatentFactors + 1)} \\
I(\scoremeans,\scorecovs) &=& \frac{1}{4}\NumLatentFactors(\NumLatentFactors+3)\sum_{\numcomponents=1}^\NumComponents\log{\NumData\weight_k} - \frac{KD}{2}\log{2} \nonumber \\ 
&& \cdots \,\, -\frac{1}{2}(2\NumLatentFactors+3)\sum_{k=1}^{K}\log{|\scorecovs_k|}  \quad . \label{eq:prior_xi_omega} 
\end{eqnarray}

@@

%The Fisher information for $\factorloads$ \todo{needs to be double checked.}
Thus the message length to encode $\factorloads$ is given by
\begin{eqnarray}
	I(\factorloads) &=& -\log\left(\frac{p(\factorloads)}{\sqrt{|\mathcal{F}(\factorloads)|}}\right) \nonumber \\
I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} \nonumber \\
	&& \cdots \,\, + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\ \label{eq:prior_L}%\nonumber \\
%	&& \cdots \todo{+\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \quad . \label{eq:prior_L}
\end{eqnarray}

%\todo{As...}% TODO: 
\vspace{10em}

Combining equations \ref{eq:prior_J}, \ref{eq:prior_K}, \ref{eq:prior_pi}, \ref{eq:prior_xi_omega}, and \ref{eq:prior_L} with equation \ref{eq:mml} leads to the full message length:
\begin{widetext}
\begin{eqnarray}
	I(\vec\Psi, \vec\data) &=& -\log\likelihood(\vec\data|\vec\Psi)
 +\frac{1}{4}\left(\NumLatentFactors + 4\right)\left(\NumLatentFactors - 1\right)\sum_{\numcomponents=1}^\NumComponents\log\weight_\numcomponents + \left(\NumComponents - \frac{1}{2}\right)\log{\NumData}
 +\frac{1}{2}\NumDimensions\log|\textrm{Cov}\left(\factorloads\transpose\right)| \nonumber \\
  && \cdots -\frac{1}{2}\left(D-J-1\right)\log|\factorloads\transpose\factorloads| + \textrm{Tr}\left(\textrm{Cov}\left(\factorloads\transpose\right)^{-1}\factorloads\transpose\factorloads\right) - \left(\NumLatentFactors + \frac{3}{2}\right)\sum_{\numcomponents=1}^\NumComponents\log|\scorecovs_\numcomponents|  -\log\Gamma\left(\NumComponents\right) - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\
&& \cdots +\frac{Q}{2}\log\kappa_q +\frac{1}{2}\left[\NumLatentFactors(\NumDimensions + 2) + \NumComponents(2-\NumData)\right]\log{2}  \quad . \nonumber \label{eq:message-length}
 \end{eqnarray}
\end{widetext}
%\vspace{10em}\clearpage
% \todo{\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}}

@@

\subsection{A toy model} \label{sec:exp-1}

Here we introduce a toy model where we use generated data to verify that
we recover the true data-generating parameters given some data, and to
ensure that the method is yielding consistent results.
We generated a data set with ${\NumData = 1}$00,000 data points, each with
$\NumDimensions = 15$ dimensions. We adopted a latent dimensional space of 
$\NumLatentFactors = 5$ factor loads such that the vector $\factorloads$ has
shape $\NumLatentFactors \times \NumDimensions$,
with $\NumComponents = 20$ clusters in the latent space. 
We generated the factor loads in the same way that we initialise the optimisation (Section~\ref{sec:initialisation}). The relative weights $\vec\weight$
are drawn from a multinomial distribution and the means of the clusters
in factor scores $\scoremeans$ are drawn from a standard normal
distribution. The off-diagonal entries in the covariance matrices in factor scores $\scorecovs$ are drawn from a gamma distribution $\scorecovs_{\numcomponents,i,i} \sim \vec\Gamma\left(1\right)$. The variance in 
each dimension $\specificvariance$ are also drawn $\specificvariance \sim \vec\Gamma\left(1\right)$.
The $\numdata$th data point (which belongs to the $\numcomponents$th cluster) is then
generated by drawing $\factorscores_{\numdata} \sim \mathcal{N}(\scoremeans_\numcomponents,\scorecovs_\numcomponents)$, projecting by the factor loads $\factorloads$, and adding variance $\specificvariance$.

%\begin{figure*}
%	\includegraphics[width=\textwidth]{experiments/exp1-latent.pdf}
%    \caption{The inferred latent scores $\factorscores$ for our toy model
%     with $\NumLatentFactors = 5$ latent factors and
%		 	 $\NumComponents = 20$ components.
%             Points are coloured by the inferred component.} 
%    \label{fig:exp1-latent}
%\end{figure*}

\begin{figure*}[t!]
	\includegraphics[width=\textwidth]{experiments/exp1-compare-all.pdf}
    \caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
    		 and specific variances $\specificvariance$ (right) compared to the 
		 	 true data generating values
		 	 for Experiment~1. The agreement is excellent.}
    \label{fig:exp1-compare}
\end{figure*}

@@

We recorded the \emph{negative} log likelihood, the BIC, and the message length (omitting constant terms) for each permutation of $\NumLatentFactors$ and $\NumComponents$.
These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.
Unsurprisingly the negative log likelihood increases with increasing numbers of latent
factors $\NumLatentFactors$ and increasing numbers of components $\NumComponents$.
The lowest BIC value and message length is found at $\NumLatentFactors = 5$
and $\NumComponents = 20$, identical to the true values.
It is clear from Figure~\ref{fig:experiment-1-gridsearch} that a combination of latent factors
and clustering in the latent space provides a better description of the (generated) data.
Adding components to the model does improves the log likelihood, even with a single latent factor,
but the addition of just \emph{one latent factor} improves the log likelihood more so than adding
\emph{twenty components}. Not much more can be said for this example because the true data generating process 
is drawn from a higher number of latent factors, but this toy model does illustrate how 
clustering in high dimensional data can be better described by latent factors with 
clustering in the lower dimensional latent space.
%The inferred clustering in latent space is shown in Figure~\ref{fig:exp1-latent},
%where each point is coloured by the inferred associated component.

@@

\begin{figure*}[t!]
	\includegraphics[width=\textwidth]{experiments/exp2-\ExperimentHash-latent-factors-and-contributions-with-targets.pdf}
	\caption{Latent factors found in Experiment~2 using the \citet{Barklem:2005}
			 data. The panels on the left show the absolute entries for each factor load. 
			 On the right we show the absolute fractional contributions to
			 each element, illustrating the loads that contribute most to each element.}
    \label{fig:exp2-factor-loads}
\end{figure*}

@@

\subsection{A toy model with data missing at random}
\label{sec:toy-model-missing-data}

Here we repeat the toy model used in the previous Section, but we discard an increasing fraction
of the data and evaluate the model's performance in the presence
of incomplete data. We considered missing data fractions from 1\% to
40\%. In each case we treated the model parameters as unknown, assumed
the missing data points were missing at random, and initialised the
model as per Section~\ref{sec:exp-1}.

\begin{figure*}[t!]
	\includegraphics[width=\textwidth]{experiments/exp-missing-data-not-at-random-inflated-40percent.png}
    \caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
    		 and specific variances $\specificvariance$ (right) compared to the 
		 	 true data generating values
		 	 for Experiment~2. Here 40\% of the data are missing at random.
			 The agreement remains excellent,
			 despite the large fraction of missing data entries.}
    \label{fig:exp1-missing-data}
\end{figure*}

In Figure~\ref{fig:exp1-missing-data} we show the results of this
experiment for our worst considered case, where 40\% of the data
entries are randomly discarded. We find that despite the high fraction
of missing entries, our estimates of the model parameters remain unbiased
in this example using a toy model. The corrections to our estimates of the
specific variances are sufficient, in that the specific variance in each
dimension is not systematically under-estimated from the true values, 
despite 40\% of the data entries to be missing.

@@

In this experiment we perform blind chemical tagging using the 
photospheric abundances released as part of the second \Galah\ 
data release \citep{Buder:2018}. This data set includes
up to 23 chemical abundances reported for 342,682 stars.
%Some abundances are not reported for a variety of 
%reasons related to astrophysics or the adopted analyses
%(e.g., a star is too hot to estimate the abundance for a given
%electronic transition, or the signal-to-noise of the spectrum is too low).
%Our method as described cannot efficiently handle missing data 
%entries for a given star, and as such we are forced to exclude stars that
%do not have a full complement of abundances. However, we also
%chose to exclude some elemental abundances from our chemical
In this example we chose to restrict ourselves to stars with a
complete set of abundance measurements for a subset of those 23 elements
(i.e., no missing data entries).
%tagging experiments. 
%For example, our experiments
For example, here we will exclude
%The element subset should not 
%with \Galah\ data do not include 
lithium and carbon abundances because the
photospheric values will vary throughout a star's lifetime \citep[e.g.,][]{Casey:2016b,Casey:2019}. This is true
to a small degree for many elements \citep[e.g.,][]{Dotter:2017},
but for the purposes of this experiment we assume that all other
photospheric abundances remain constant throughout a star's
lifetime.

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-latent-factors-and-contributions-with-targets.pdf}
	\caption{Latent factors inferred from 1,072 stars in \Galah\
			 \citep[][thick lines]{Buder:2018} with 17 abundance measurements. Left panels show the absolute entries for each
			 factor load, where the thin lines indicate the target latent factors (see Section~\ref{sec:exp4}). On the right we show the absolute fractional contributions
			 to each element, ordered by the loads that contribute most.}
    \label{fig:exp3-factor-loads}
\end{figure*}

@@
